A Test of the Frequency Independence Assumption of Power System Parameters used in Geomagnetically Induced Current Estimates

### 2019-03-07 paper.tex
Updates from Overleaf

\title{A Test of the Frequency Independence Assumption of the Power System Parameters used in Geomagnetically Induced Current Calculations}

@@

\begin{keypoints}
\item Power system parameters derived from a set of GIC and geoelectric field measurements are frequency dependent.
\item A GIC prediction model with frequency dependence significantly outperforms a frequency independent model.
\item A model derived using GIC/geomagnetic field data provides better predictions than a GIC/geoelectric field-based model.
\end{keypoints}

\begin{abstract}
A common assumption used in estimating geomagnetically induced currents (GICs) in a power system given a nearby direct measurement or indirect estimate (via a transfer function with input of the geomagnetic field) of the horizontal geoelectric field components $E_x$ and $E_y$ on Earth's surface is that the system is resistive. That is, the approximation $GIC(t)$  $=$ $a_oE_x(t) + b_oE_y(t)$ (Model 1) is used, where $a_o$ and $b_o$ are frequency independent network parameters.  We provide the first test of this assumption using GIC measurements made in a 187 kV transformer connected to a $\sim$100~km power line in Memanbetsu, Japan and geoelectric field measurements made at the Memanbetsu Magnetic Observatory $\sim$9~km away.  We compute frequency dependent network parameters $a(\omega)$ and $b(\omega)$ in Model 2, $GIC(\omega)$ $=$ $a(\omega)E_x(\omega) + b(\omega)E_y(\omega)$, and find that they are frequency dependent and that this model provides a significantly better representation of measured GIC.
\end{abstract}

@@

Geomagnetically induced currents (GICs) are electric currents in conducting systems due to electric fields induced near Earth's surface. These fields are a result of time variations in ionospheric currents on timescales on the order of hours \citep{Ohtani2000} and the movement of Earth's surface relative to slow-varying current systems in Earth's ionosphere that are near stationary relative to the Sun \citep{Stening2013}. Of particular interest are currents induced in electric power systems as they can lead to system degradation, disruption, and failure \citep{Albertson1993,NERC2012}. Accurate estimation of GICs given direct measurements of the electric field (or estimates of the electric field based on the geomagnetic field) is important for power system design, retrospective analysis, and mitigation of the impacts of space weather on power systems \citep{Molinski2002,Thomson2010,NERC2012,Gaunt2014}.

A common assumption made in estimates of GIC using either measured or estimated values of the horizontal geoelectric field at Earth's surface, $\mathbf{E}$, is that the power system is resistive or quasi-DC in the sense that the relationship $GIC(t) = a_oE_x(t) + b_oE_y(t)$ holds \citep{Albertson1981,Lehtinen1985}. In this case, estimates of the coefficients $a_o$ and $b_o$ can be easily made from the data using linear regression, when contemporaneous measurements of GIC and $\mathbf{E}$ are available, or by using information about the connectivity of the power lines and values of the conductor and transformer resistances using DC circuit methods \citep[e.g.][]{Boteler2014a,Boteler2014b}. This quasi-DC assumption has been used or stated in many works, for example, \citet{Pulkkinen2007,Wik2008,Pulkkinen2010,Ngwira2011,Horton2012,Viljanen2012,Overbye2012,Marshall2013,Liu2014,Zheng2014,Watari2015,Bonner2017}.

@@

In this work, we use a unique dataset in which measurements $\mathbf{E}$ and the surface horizontal geomagnetic field, $\mathbf{B}$, were made near a site where GICs were being measured in an electric power system.  We first estimate the system parameters $a_o$ and $b_o$ from data using conventional methods, which assume that they are frequency independent. Then, we use a model in which the system parameters are frequency dependent, compute the frequency dependence, and compare this model with the traditional frequency independent model.

@@

The 1-second-cadence GIC data used span a total of 35 days across 10 intervals that start and end at midnight Japan Standard Time (JST): 2006-04-03 -- 2006-04-10, 2006-07-25 -- 2006-07-29, 2006-08-05 -- 2006-08-08, 2006-08-19 -- 2006-08-21, 2006-11-07 -- 2006-11-12, 2006-11-28 -- 2006-11-29, 2006-12-01, 2006-12-14 -- 2006-12-15, 2007-11-18 -- 2007-11-20, and 2007-11-22 -- 2007-11-22. These data have been presented previously in \citet{Watari2009} and \cite{Watari2015}. These time intervals cover those for which GIC data was made available with the exception of three days for which there were obvious problems with either the electric field or GIC measurements. The GIC dataset includes 1~s raw data and 1~Hz low-pass filtered raw measurements. The results presented here are for the 1~Hz low-pass filtered GIC values

One-second-cadence $\mathbf{E}$ and $\mathbf{B}$ measurements a from Memambetsu Magnetic Observatory were obtained from the Japan Meteorological Agency data portal.

@@

Spikes in the electric field measurements were assumed to be unphysical when the absolute value of a component $i$ of the electric field changed by more than 1 mV/km at time $t_s$, and the values $E_i(t_s-2)$ through $E_i(t_s+4)$ were replaced using linear interpolation of the values at $E_i(t_s-3)$ and $E_i(t_s + 5)$. In the sample shown in Figure~\ref{sample}, 21 spikes were identified in $E_x$ and zero in $E_y$. For all of the time intervals considered, the fraction of interpolated $\mathbf{E}$ values ranged from 0--0.1\%.

The only modification made to the magnetic field measurements was the replacement of 36 data points in all components of the magnetic field on 2006-08-06 with linearly interpolated values. The 36 values for $B_x$ and $B_y$ were the same and much larger than the surrounding values.

@@

\noindent
the matrix equation $\mathbf{GIC} = \underline{\mathbf{E}}\cdot\mathbf{p}$ is created, where $\mathbf{GIC}$ is a 864000x1 matrix of 1 day of observations with zero mean, $\mathbf{p} = [a_o,b_o]^T$, and $\underline{\mathbf{E}}$ is a 86400x2 matrix with rows of $[E_x(t), E_y(t)]$ observations with each column having zero mean. The values of $\mathbf{p}$ are then obtained using a standard linear regression solver. The values of $a_o$ and $b_o$ determined in this way correspond to those that minimize the sum-of-squares of $GIC(t)-G_o(t)$ over a 1-day interval. \citep[][provided the mathematically equivalent closed-form equations.]{Pulkkinen2007}. The averages $\overline{a}_o$ and $\overline{b}_o$ are the average of the 35 $a_o$ and $b_o$ values computed using the 35 unique 1-day window segments of available data.

@@

\begin{linenomath*}
\begin{equation}
G_E(\omega) = a(\omega)E_x(\omega) + b(\omega)E_y(\omega)
\label{model2}
\end{equation}
\end{linenomath*}
\noindent
and are obtained using contemporaneous measurements of GIC and $\mathbf{E}$ using the method described later in this section. Note that fourier transformed variables are indicated by explicitly listing their dependence on $\omega$.

In Model~3, the parameters $a(\omega)$ and $b(\omega)$ from Model~2 are used, but with an electric field $\mathbf{E}'$ computed using the measured magnetic field and a transfer function $\boldsymbol{Z}$: $\mathbf{E}'(\omega) \equiv \boldsymbol{Z}(\omega)\mathbf{B}(\omega)/\mu_o$:

@@

\noindent
where, explicitly, the electric field components are $\mu_oE'_x$ = $Z_{xx}B_x + Z_{xy}B_y$ and $\mu_oE'_y = Z_{yx}B_x + Z_{yy}B_y$. Substitution gives

\begin{linenomath*}
\begin{equation*}
G_{E'}(\omega) = \big[a(\omega)Z_{xx}(\omega) + b(\omega)Z_{yx}(\omega) \big] B_x(\omega)/\mu_o + \big[ a(\omega)Z_{xy}(\omega) + b(\omega)Z_{yy}(\omega) \big] B_y(\omega)/\mu_o\,.
\end{equation*}
\end{linenomath*}

That is, instead of using the measured electric field directly as in Model 2, a transfer function that relates $\mathbf{E}$ to $\mathbf{B}$ is used to provide an estimate of the electric field based on the magnetic field. In this work, $\boldsymbol{Z}$ was calculated using the method described later in this section.

@@

\noindent
In this model, the transfer function components $z_x$ and $z_y$ are determined directly from measurements of GIC and $\mathbf{B}$ and the model is expected to produce predictions of GICs that are similar to that obtained from Model 3.

Estimates of $a(\omega)$ and $b(\omega)$ are made at evaluation frequencies, $\omega_e \equiv 2\pi f_e$, by performing a least squares regression using Equation~\ref{model2} and the set $G_E(\omega)$ and $\mathbf{E}(\omega)$ values near $f_e$ from the fourier transform of 24 hours of 1-s cadence data. The values of $f_e$ start at $0.25$ Hz and decrease by a factor of $\sqrt{2}$; if $f_e$ value is not a multiple of $1/86400$, the nearest multiple is chosen. The regression for $f_e$ uses all $f$ values between the next higher and lower value of $f_e$.

@@

Transfer function parameters were computed for each of the 35 unique 1-day window segments of available data. The average transfer function magnitude was obtained by averaging these 35 transfer function magnitudes; the average transfer function phase was obtained using the averaged real and imaginary components, respectively, of the 35 transfer functions. The average impulse response functions were obtained by averaging the 35 impulse response functions.

There are many statistical methods and variations thereof that have been developed for estimating transfer functions in the magnetotelluric (MT) literature that use additional steps and processing \citep{Egbert1986,Chave1987,Chave1989,Jones1989,Larsen1996,Egbert1997,Eisel2001,Chave2004,Chave2012,Chave2017} beyond the ones used that follow the basic method in \cite{Sims1971} and \cite{Simpson2005}. The objective of these works is typically the estimation of a robust and low bias transfer function. In the results presented, we did not use a robust statistical method or attempt to correct for bias in estimates as our primary interest is comparing the ability of different models to predict the data (in a mean-squared-error sense) rather than estimating the model parameters. As discussed and evaluated in~\cite{Weigel2017}, a model developed using robust and bias correction methods will not necessarily be an optimal prediction model (when the sum-of-squares error is used as a prediction metric).

However, several additional processing methods and steps used in the MT literature were tested individually by modifying the basic least-squares method used to determine if there was a change in the computed transfer functions and, most relevant for this work, the out-of-sample prediction performance of each model. When using these additional steps, we find that the conclusions made regarding the differences each model's prediction ability do not change, the differences in each model's prediction performance is within the calculated uncertainty of the used method, and the transfer function values are generally within the error bars of the used method when the signal to noise ratio is above $\sim$3 (corresponding to a period of $\sim 30$~s). The experiments in this regard are discussed in the Appendix.

@@

The out-of-sample prediction metrics for each model on a given day was determined by computing the average model parameters using the remaining 34 days of data, generating a prediction time series on each of the 34 days using the average model parameters, and then computing the prediction metrics for each of the 34 days. This process was repeated to create a set of 35 out-of-sample prediction metrics for each model.

The performance of each model in predicting GIC was assessed using three metrics: the prediction efficiency, PE, the mean-squared-error (MSE) ratio, and the signal-to-noise ratio. 
The prediction efficiency, PE, is the ``Case I'' skill score described by \cite{Murphy1988} and is defined as PE = 1 - variance(Prediction Error)/variance(Predicted); it is a measure of the skill of the model relative to the variance in the data that are predicted, and it represents the fraction of the variance in the predicted data that predicted by the model.

The average PE and MSE ratio was computed by averaging the 35 out-of-sample PE and MSE ratios; the average signal-to-noise ratio at a given frequency is the average of the 35  out-of-sample signal-to-noise ratios at that frequency. The 95\% confidence interval, CI, on the PE and MSE ratio was calculated using 1000 bootstrap samples of the 35 prediction metrics. The typical difference between the limits when the confidence intervals were computed assuming the values were Gaussian-distributed was $\sim$2\%.

@@

Figure~\ref{predictions} shows a 1-day interval for selected because the value of the prediction efficiency for each model was in the same order as that shown in Table~1, i.e., Model 1 has the lowest PE and Model 4 has the lowest.

The primary conclusions of this work follow from the results shown in Table~1: the frequency dependent model, Model 2, is significantly better at predicting GIC than the frequency independent model, Model 1. Model 3, which uses an estimate of the electric field based on horizontal magnetic field measurements provides better estimates of GIC than Model 2. Model 4, in which a transfer function that connects GIC to $\mathbf{B}$ was derived directly from the data, provides the best predictions of GIC.

@@

Figure~\ref{SN} shows the average signal-to-noise ratios as a function of period for the four models. The error bars represent $\pm$ one standard error of the 35 SN ratios at each period. At each evaluation frequency the signal-to-noise ratio is is the ratio of the average of the 35 segment PSDs of GIC to the average of the 35 segment PSDs of the prediction error. The PSD for each of the segments at a given evaluation frequency was obtained by averaging the raw PSD using values between the next higher and lower evaluation frequency.

Consistent with the results in Table~1, The SN ratio for Model 1 is lower than that for Models 2-4 at all frequencies.

Below $600$~s (10~min), Model 2 has the largest SN ratio. This result is somewhat visible in the sample interval shown in Figure~\ref{predictions} where the high frequency fluctuations in the prediction error for Model 2 are smaller than that for Models 3 and 4.

Below $600$~s (10~min), the predictive ability of each model is in the ordering that may be expected based on the data used to derive the model: Model 2, which uses direct measurements of the electric field, has a higher SN ratio than Model 3, which uses an indirect estimate of the electric field. Model 4 has a slightly higher SN than Model 3, and this is expected because they both use the same input of the magnetic field, but Model 3 has an additional source of uncertainty from the parameters in $\mathbf{Z}$. Above $600$~s, the ordering of the SN is not consistent with this expectation and is possibly explained by two factors (1) the measured GIC is based on an integrated electric field and the electric field used in Model 2 is a point measurement - the result is that implicit in the model is an assumption that the electric field is uniform over the length of the line on which GIC is measured and (2) the error associated with this assumption has less of an influence on the errors in a model that uses the point measurements of the magnetic field.

@@

The average parameters and 95\% CI for Model 1 are $a_o = $ $73 [62,64]$ A/(V/km) and $b_o =$ $-72.0 \pm [-98,-45]$ A/(V/km) and the distribution of the 35 parameters calculated using each 1-day interval is shown in Figure~\ref{histogram}. (Results are similar if 1-minute averages are used; \cite{Watari2015} found $a_o=38.1$ A/(V/km) and $b_o=-7.4$ A/(V/km) for the interval 2006-12-14--2006-12-15 JST time interval).

@@

The average parameters for Models 2 and 4 are shown in Figure~\ref{H}--\ref{Phi}. The error bars at each time step or period is the standard error of the 35 values used to compute the average. The transfer function magnitudes were calculated by linear interpolation of the transfer function magnitudes at the evaluation frequencies (instead of linearly interpolating the real and imaginary components and then computing the magnitude).

Figure~\ref{H} shows that the timescale of variations of an impulse in $\mathbf{E}$ for Model 2 and $\mathbf{B}$ for Model 4 are similar and are different from zero at more than one time value (the impulse response for Model 3 is similar to that of Model 2 and is not shown). By definition, the impulse response for Model 1 is zero except at $t=0$. Note that the sign of the impulse response at $t=0$ is opposite for Model 1 and Model 2 for the $E_y$ coefficient - at $t=0$, $b_o < 0$ while $h_b(t=0)>0$. An explanation for the acausality of the impulse response is given in \cite{Kelbert2017}.

Figure~\ref{Z} shows that the frequency domain transfer functions for Model 2 vary by an order of magnitude on timescales of 60 seconds to 12 hours (that for Model 1 is constant by definition). In Model~2, GIC more sensitive to $E_y$ due to $b(\omega) > a(\omega)$, but note that $E_y$ has less variability than $E_x$ - the variance in $E_x$ is $\sim$5x larger than $E_y$ over the entire dataset (this is also visible in Figure~\ref{sample}). The ratio of $|\overline{b}_o/|/|\overline{a}_o|$ is approximately 1 for Model~1 and this can be compared with $|b(\omega)|/|a(\omega)|$ calculated using the lines for $|a(\omega)|$ and $|b(\omega)|$ shown in Figure~\ref{Z}. The ratio varies from 2.7 at $T=30$~s to 1.6 at $T=12$~h. At $T=12$~h, the highest period considered, the phases of $a(\omega)$ and $b(\omega)$ shown in Figure~\ref{Phi} differ by 180 degrees and the signs are consistent with $\overline{a}_o$ and $\overline{b}_o$.

Figure~\ref{Phi} shows the transfer function phases. By definition, that for Model 1 is zero. The phases for $z_x(\omega)$ and $b(\omega)$ are offset by 120 degrees while the phases for $a(\omega)$ and $b(\omega)$ are significantly different. \todo{Why?} At a period of $60$~s, the parameters in Equation~\ref{model2} are most consistent with the assumptions of frequency independent system parameters of Model~1. The phases of $a(\omega)$ and $b(\omega)$ are near zero and the $b(\omega)$ curve is the flattest.

@@

There are at least three possible explanations of the frequency dependence of the data-derived system parameters: (1) the quasi-DC approximation is invalid, (2) invalid model assumptions, and (3) artifacts of the measurement process.

(1) When the quasi-DC approximation is used, ~\cite{Albertson1981} is usually cited; the justification for the quasi-DC approximation is given the Appendix, where it is stated that ``An analysis based on Carson's equations reveals that transmission lines can be modeled for GIC determination by using their positive sequence resistance values with a small correction factor to account for skin and proximity effects. The correction factor varies from 0.95 to 1.0 depending on conductor size.'' \cite{Lehtinen1985} notes justifies the use of the quasi-DC approximation because GIC amplitudes are significant at frequencies that are much lower than the 50~Hz power line frequency. Carson's equations are equations for the frequency-dependent impedance associated with a wire above a conducting plane. The justification of \cite{Lehtinen1985} is essentially a statement that the frequency dependent component of Carson's impedance is proportional to frequency, the frequency dependent impedance of a power line is small at 50~Hz and so it should be even smaller at GIC frequencies of < 1~Hz. \todo{Need a reference here to 50~Hz values} Thus it appears that the quasi-DC approximation is justified.

>>>>

\title{A Test of the Frequency Independence Assumption of Power System Parameters used in Geomagnetically Induced Current Estimates}

@@

\begin{keypoints}
\item Power system parameters derived from a time series of GIC and geoelectric field measurements are frequency dependent.
\item A GIC prediction model with frequency dependence significantly outperforms a frequency independent model.
\item A model derived using GIC/geomagnetic field data provides better predictions than a GIC/geoelectric field-based model.
\end{keypoints}

\begin{abstract}
A common assumption used in estimating geomagnetically induced currents (GICs) in a power system given a nearby direct measurement or indirect estimate (via a transfer function with input of the geomagnetic field) of the horizontal geoelectric field components $E_x$ and $E_y$ on Earth's surface is that the system is resistive. That is, the approximation $GIC(t)$  $=$ $a_oE_x(t) + b_oE_y(t)$ (Model 1) is used, where $a_o$ and $b_o$ are frequency independent network parameters.  We provide the first test of this assumption using GIC measurements made in a 187 kV transformer connected to a $\sim$100~km power line in Memanbetsu, Japan and geoelectric field measurements made at the Memanbetsu Magnetic Observatory $\sim$9~km away.  We compute frequency dependent network parameters $a(\omega)$ and $b(\omega)$ from data in Model 2, $GIC(\omega)$ $=$ $a(\omega)E_x(\omega) + b(\omega)E_y(\omega)$, and find that they are frequency dependent and that this model provides a significantly better estimates of the measured GIC. A third model, which uses the electric field estimated from a transfer function convolved with the geomagnetic field, provides significantly better estimates to Model~2. It is suggested that the measurement-derived frequency dependence of the system parameters is explained by spatial variations in the spectra of the geoelectric field rather than by an actual frequency dependence of the system parameters.
\end{abstract}

@@

Geomagnetically induced currents (GICs) are electric currents in conducting systems due to electric fields near Earth's surface. Time-varying electric and magnetic fields at Earth's surface are driven by time variations in ionospheric currents on time scales on the order of hours \citep{Ohtani2000} and the movement of Earth's surface relative to slow-varying current systems in Earth's ionosphere that are near stationary relative to the Sun \citep{Stening2013}. Of particular interest are currents induced in electric power systems as they can lead to system degradation, disruption, and failure \citep{Albertson1993,NERC2012}. Accurate estimation of GICs magnitudes is important for power system design, retrospective analysis, and mitigation of the impacts of space weather on power systems \citep{Molinski2002,Thomson2010,NERC2012,Gaunt2014}.

A common assumption made in estimates of GIC using either measured or estimated values of the horizontal geoelectric field at Earth's surface, $\mathbf{E}$, is that the power system is resistive or quasi-DC in the sense that the relationship $GIC(t) = a_oE_x(t) + b_oE_y(t)$ holds \citep{Albertson1981,Lehtinen1985}. In this case, estimates of the coefficients $a_o$ and $b_o$ can be easily made from the data using linear regression, when contemporaneous measurements of GIC and $\mathbf{E}$ are available, or by using information about the connectivity of the power lines and values of the conductor and transformer resistances using DC circuit methods \citep[e.g.][]{Boteler2014a,Boteler2014b}. This quasi-DC assumption has been used or stated in many GIC-related works, for example, \citet{Pulkkinen2007,Wik2008,Pulkkinen2010,Ngwira2011,Horton2012,Viljanen2012,Overbye2012,Marshall2013,Liu2014,Zheng2014,Watari2015,Bonner2017}.

@@

In this work, we use a unique dataset in which measurements $\mathbf{E}$ and the surface horizontal geomagnetic field, $\mathbf{B}$, were made near a site where GICs were being measured in an electric power system.  We first estimate the system parameters $a_o$ and $b_o$ from data using conventional methods, which assume that they are frequency independent. Then, we use a model in which the system parameters are allowed to be frequency dependent, compute the frequency dependence, and compare this model with the traditional frequency independent model.

@@

The 1-second-cadence GIC data used span a total of 35 days across 10 continuous intervals that start and end at midnight Japan Standard Time (JST): 2006-04-03 -- 2006-04-10, 2006-07-25 -- 2006-07-29, 2006-08-05 -- 2006-08-08, 2006-08-19 -- 2006-08-21, 2006-11-07 -- 2006-11-12, 2006-11-28 -- 2006-11-29, 2006-12-01, 2006-12-14 -- 2006-12-15, 2007-11-18 -- 2007-11-20, and 2007-11-22 -- 2007-11-22. These data have been presented previously in \citet{Watari2009} and \cite{Watari2015}. These time intervals are those for which GIC data was made available with the exception of three days for which there were obvious problems with either the electric field or GIC measurements. The GIC dataset includes 1~s raw data and 1~Hz low-pass-filtered raw measurements. The results presented here are for the 1~Hz low-pass-filtered GIC values

One-second-cadence $\mathbf{E}$ and $\mathbf{B}$ measurements from Memambetsu Magnetic Observatory were obtained from the Japan Meteorological Agency data portal.

@@

Spikes in the electric field measurements were assumed to be unphysical when the absolute value of a component $i$ of the electric field changed by more than 1~mV/km at time $t_s$, and the values $E_i(t_s-2)$ through $E_i(t_s+4)$ were replaced using linear interpolation of the values at $E_i(t_s-3)$ and $E_i(t_s + 5)$. In the sample shown in Figure~\ref{sample}, 21 spikes were identified in $E_x$ and zero in $E_y$. For all of the time intervals considered, the fraction of interpolated $\mathbf{E}$ values ranged from 0--0.1\%.

The only modification made to the magnetic field measurements was the replacement of 36 data points in all components on 2006-08-06 with linearly interpolated values. The 36 values for $B_x$ and $B_y$ were the same and much larger than the surrounding values.

@@

\noindent
the matrix equation $\underline{\mathbf{GIC}} = \underline{\mathbf{E}}\cdot\mathbf{p}$ is created, where $\underline{\mathbf{GIC}}$ is a zero mean 864000x1 matrix of 1 day observations with rows of $GIC(t)$; $\mathbf{p} = [a_o,b_o]^T$, and $\underline{\mathbf{E}}$ is a 86400x2 matrix with rows of $[E_x(t), E_y(t)]$ observations with each column having zero mean. The values of $\mathbf{p}$ are then obtained using linear regression. The values of $a_o$ and $b_o$ determined in this way correspond to those that minimize the sum-of-squares of $GIC(t)-G_o(t)$ over a 1-day interval. \citep[][provided the mathematically equivalent closed-form equations.]{Pulkkinen2007}. The notation $\overline{a}_o$ and $\overline{b}_o$ is used for the average of the $a_o$ and $b_o$ values computed using the 35 unique 1-day window segments of available data.

@@

\begin{linenomath*}
\begin{equation}
G_E(\omega) = a(\omega)E_x(\omega) + b(\omega)E_y(\omega)
\label{model2}
\end{equation}
\end{linenomath*}
\noindent
and are computed using contemporaneous measurements of $GIC(t)$ and $\mathbf{E}(t)$ using the method described later in this section. Note that fourier transformed variables are indicated by explicitly listing their dependence on $\omega$.

In Model~3, the parameters $a(\omega)$ and $b(\omega)$ from Model~2 are used, but with an electric field $\mathbf{E}'$ computed using the measured magnetic field and a transfer function $\boldsymbol{Z}$: $\mathbf{E}'(\omega) \equiv \boldsymbol{Z}(\omega)\mathbf{B}(\omega)$:

@@

\noindent
where, explicitly, the electric field components are $E'_x$ = $Z_{xx}B_x + Z_{xy}B_y$ and $E'_y = Z_{yx}B_x + Z_{yy}B_y$. 
That is, instead of using the measured electric field directly as in Model 2, a transfer function that relates $\mathbf{E}$ to $\mathbf{B}$ is used to provide an estimate of the electric field based on the magnetic field. In this work, $\boldsymbol{Z}$ was calculated using the method described later in this section.

%Substitution gives

%\begin{linenomath*}
%\begin{equation*}
%G_{E'}(\omega) = %\big[a(\omega)Z_{xx}(\omega) + b(\omega)Z_{yx}(\omega) \big] B_x(\omega) + \big[ a(\omega)Z_{xy}(\omega) + b(\omega)Z_{yy}(\omega) \big] B_y(\omega)\,.
%\end{equation*}
%\end{linenomath*}

@@

\noindent
In this model, the transfer function components $z_x$ and $z_y$ are determined directly from measurements of $GIC(t)$ and $\mathbf{B}(t)$, and the model is expected to produce predictions of $GIC(t)$ that are similar to that obtained from Model 3 because they both use the same input variable.

Estimates of $a(\omega)$ and $b(\omega)$ are made at evaluation frequencies, $\omega_e \equiv 2\pi f_e$, by performing a least squares regression using Equation~\ref{model2} and the set $G_E(\omega)$ and $\mathbf{E}(\omega)$ values near $f_e$ from the fourier transform of 24 hours of 1-s cadence data. The values of $f_e$ start at $0.25$ Hz and decrease by a factor of $\sqrt{2}$; if $f_e$ value is not a multiple of $1/86400$, the nearest multiple is chosen. The regression for $f_e$ uses all $f$ values in a window with a width of twice the separation from the next lower value of $f_e$ and the given $f_e$.

@@

Transfer function parameters were computed for each of the 35 1-day segments of available data. The average transfer function magnitude and phase was obtained by calculating their values each segment and then averaging the segments values.

There are many statistical methods and variations thereof that have been developed for estimating transfer functions in the magnetotelluric (MT) literature that use additional steps and processing \citep{Egbert1986,Chave1987,Chave1989,Jones1989,Larsen1996,Egbert1997,Eisel2001,Chave2004,Chave2012,Chave2017} beyond the ones used in this work that follow the basic method used in \cite{Sims1971} and \cite{Simpson2005}. Several additional processing methods and steps beyond the basic method that have been used in the MT literature were tested individually by modifying the basic least-squares method used to determine if there was a change in the computed transfer functions and, most relevant for this work, the out-of-sample prediction performance of each model. When using these additional steps, we find that the conclusions made regarding the differences each model's prediction ability do not change, the differences in each model's prediction performance is within the calculated uncertainty of the used method, and the transfer function values are generally within the error bars of the basic method used when the signal to noise ratio is above $\sim$3 (corresponding to a period of $\sim 30$~s). The experiments in this regard are discussed in the Appendix.

%The objective of these works is typically the estimation of a robust and low bias transfer function.

%In the results presented, we did not use a robust statistical method or attempt to correct for bias in estimates as our primary interest is comparing the ability of different models to predict the data (in a mean-squared-error sense) rather than estimating the model parameters. As discussed and evaluated in~\cite{Weigel2017}, a model developed using robust and bias correction methods will not necessarily be an optimal prediction model (when the sum-of-squares error is used as a prediction metric).

@@

The out-of-sample prediction metrics for each model on day $D$ were determined by computing the average model parameters using the remaining 34 days of data, using these averaged parameters to generate a prediction time series on day $D$. This process was repeated to create a set of 35 out-of-sample prediction metrics for each model.

The performance of each model in predicting GIC was assessed using three metrics: the prediction efficiency, PE, the mean-squared-error (MSE) ratio, and the signal-to-noise ratio.

The prediction efficiency, PE, is the ``Case I'' skill score described by \cite{Murphy1988} and is defined as PE = 1 - variance(Prediction Error)/variance(Predicted); it is a measure of the skill of the model relative to the variance in the data that are predicted, and it represents the fraction of the variance in the data that predicted by the model.

The average PE and MSE ratios were computed by averaging the 35 out-of-sample PE and MSE ratios; the average signal-to-noise ratio at a given frequency is the average of the out-of-sample signal-to-noise ratios at that frequency. The 95\% confidence interval, CI, on the PE and MSE ratio was calculated using 1000 bootstrap samples of the out-of-sample prediction metrics. The typical difference between the limits when the confidence intervals were computed assuming the values were Gaussian-distributed was $\sim$2\%.

@@

Figure~\ref{predictions} shows the predictions for each model in a 1-day interval selected because the value of the prediction efficiency for each model was in the same order as that shown in Table~1, i.e., Model 1 has the lowest PE and Model 4 has the lowest.

The primary conclusions of this work follow from the results shown in Table~1: the frequency dependent model, Model~2, is significantly better at predicting GIC than the frequency independent model, Model~1; Model~3, which uses an estimate of the electric field based on magnetic field measurements, provides significantly better estimates of GIC than Model~2; Model~4, in which a transfer function that connects GIC to $\mathbf{B}$ was derived directly from the data, provides significantly better estimates of GIC than Model~3.

In terms of statistical significance, using $10^5$ bootstrap samples, the null hypothesis that the PE of Model~2 is equal to that of Model~1 can be rejected at a significance level of less than $10^{-5}$, with the same result for Model~3 vs. Model~2. The null hypothesis that the PE of Model~4 is equal to that of Model~3 can be rejected at a significance level of $10^{-2}$.

@@

Figure~\ref{SN} shows the average signal-to-noise (i.e., signal to prediction error) ratios as a function of period for the four models. At each evaluation frequency, the SN ratio is the ratio of the average of the SN ratios for each of the 35 segments and the error bars represent 95\% CIs based on 1000 bootstrap samples of the segment SN ratios at each period. The spectra for the signal and prediction error for each of the segments at a given evaluation frequency was obtained by averaging the raw spectra using values between the next higher and lower evaluation frequency.

Consistent with the results in Table~1, The SN ratio for Model~1 is lower than that for Models~2-4 at all frequencies. However, ordering of the SN ratio for Models~2-4 is dependent on period.

Below $600$~s (10~min), Model~2 has the largest SN ratio. This result is somewhat visible in the sample interval shown in Figure~\ref{predictions} where the high frequency fluctuations in the prediction error for Model~2 are the smallest among the models. For these periods, the predictive ability of each model is in the order that may be expected based on the data used to derive the model: Model~2, which uses direct measurements of the electric field, has a higher SN ratio than Model~3, which uses an indirect estimate of the electric field. Model~4 has a slightly higher SN than Model~3, and this is expected because they both use the same input of the magnetic field, but Model~3 has an additional source of uncertainty from the parameters in $\mathbf{Z}$. Above $600$~s, the ordering of the SN is not consistent with these expectations and is possibly explained by two factors (1) the measured GIC is based on an integrated electric field along the power line, and the electric field used in Model~2 is a point measurement - the result is that implicit in the model is an assumption that the electric field is uniform over the length of the line on which GIC is measured and (2) the error associated with this assumption has less of an influence on the errors in a model that uses the point measurements of the magnetic field. This issue is addressed in the following section.

@@

The average parameters and 95\% CI for Model 1 are $a_o = $ $73 [62,64]$ A/(V/km) and $b_o =$ $-72.0 \pm [-98,-45]$ A/(V/km) and the distribution of the parameters for the segments is shown in Figure~\ref{histogram}. (Results are similar if 1-minute averages are used; \cite{Watari2015} found $a_o=38.1$ A/(V/km) and $b_o=-7.4$ A/(V/km) for the interval 2006-12-14--2006-12-15 JST time interval).

@@

The average parameters for Models~2 and~4 are shown in Figure~\ref{H}--\ref{Phi}. The error bars at each time step or period is the standard error of the 35 values used to compute the average. The transfer function magnitudes were calculated by linear interpolation of the transfer function magnitudes at the evaluation frequencies (instead of linearly interpolating the real and imaginary components and then computing the magnitude).

Figure~\ref{H} shows that the timescale of variations of an impulse in $\mathbf{E}$ for Model~2 and $\mathbf{B}$ for Model~4 are similar and are different from zero at more than one time value (the impulse response for Model~3 is similar to that of Model~2 and is not shown). By definition, the impulse response for Model~1 is zero except at $t=0$. Note that the sign of the impulse response at $t=0$ is opposite for Model~1 and Model~2 for the $E_y$ coefficient (at $t=0$, $b_o < 0$ while $h_b(t=0)>0$). An explanation for the acausality of the impulse response is given in \cite{Egbert1992,Kelbert2017}.

Figure~\ref{Z} shows that the frequency domain transfer functions for Model~2 vary by an order of magnitude on timescales of 60~s to 12~h (that for Model~1 is constant by definition). For this system, GIC more sensitive to $E_y$ due to $b(\omega) > a(\omega)$, but note that $E_y$ has less variability than $E_x$ - the variance in $E_x$ is $\sim$5x larger than $E_y$ over the entire dataset (this is also visible in Figure~\ref{sample}). The ratio of $|\overline{b}_o/|/|\overline{a}_o|$ is approximately 1 for Model~1 and this can be compared with $|b(\omega)|/|a(\omega)|$ calculated using the lines for $|a(\omega)|$ and $|b(\omega)|$ shown in Figure~\ref{Z}. The ratio varies from 2.7 at $T=30$~s to 1.6 at $T=12$~h. At $T=12$~h, the highest period considered, the phases of $a(\omega)$ and $b(\omega)$ shown in Figure~\ref{Phi} differ by 180 degrees and the signs are consistent with $\overline{a}_o$ and $\overline{b}_o$.

Figure~\ref{Phi} shows the transfer function phases (that for Model~1 is 0$^o$ for positive parameters and 180$^o$ for negative parameters by definition). The phases for $z_x(\omega)$ and $b(\omega)$ are offset by 120$^o$ while the phases for $a(\omega)$ and $b(\omega)$ are significantly different.

@@

There are at least three possible explanations for the frequency dependence of the data-derived system parameters: (1) the quasi-DC approximation is invalid, (2) it is an artifact of the measurement process, and (3) the electric field is not spatially and spectrally constant across the region of the power system.
